---
title: "Finding Admissible Two-Stage Trial Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Admissible Two-Stage Trial Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MyTools)
```

## Overview

The **twostage** module of **MyTools** enumerates admissible two-stage
single-arm trial designs for a binary endpoint. It is aimed at early-phase
oncology studies where a Simon-like design is needed but the full Pareto
frontier — not just the minimax or optimal design — is of interest.

The main entry point is `twostage_admissible_designs()`, which takes the
four clinical constraints and returns all Pareto-optimal designs in a
single call.

## Design parameters

Each design is defined by five integers $(n_1, n, r_1, e_1, r)$:

| Parameter | Role |
|-----------|------|
| $n_1$ | Stage-1 sample size |
| $n$ | Total sample size ($n_2 = n - n_1$ enrolled in stage 2) |
| $r_1$ | Futility boundary: stop early if $X_1 \le r_1$ ($-1$ = no futility stop) |
| $e_1$ | Interim efficacy boundary: declare success if $X_1 \ge e_1$ ($n_1+1$ = no interim efficacy stop) |
| $r$ | Final success threshold: declare success if $X_1 + X_2 \ge r$ |

By default no $e_1 \ge r$ constraint is applied. Pass `non_binding = TRUE`
to restrict to designs where $e_1 \ge r$, guaranteeing that once interim
efficacy is declared no stage-2 outcome can overturn it (worst case $X_2 = 0$
still gives $X_1 + X_2 \ge e_1 \ge r$). This is appropriate when the interim
efficacy stopping rule is **non-binding** — i.e. the trial may continue to
stage 2 even after declaring interim success.

## Finding admissible designs

Suppose the null response rate is $p_0 = 0.10$ and the target alternative is
$p_1 = 0.30$. We want at most 5% type I error and at least 80% power, with
total sample size no larger than 40.

```{r constraints}
p0    <- 0.10
p1    <- 0.30
alpha <- 0.05
power <- 0.80
n_max <- 40L
```

`twostage_admissible_designs()` performs an exhaustive search and returns
the Pareto-optimal designs on three criteria simultaneously: maximum sample
size ($n$), expected sample size under the null ($EN_0$), and expected
sample size under the alternative ($EN_1$). Four designs of interest are
flagged in the `design_type` column: the **minimax** (minimum $n$), the
**optimal null** (minimum $EN_0$), the **optimal alt** (minimum $EN_1$), and
the **optimal mean** (minimum $\tfrac{1}{2}(EN_0 + EN_1)$) designs. A design
meeting several criteria carries a comma-joined label.

```{r admissible}
result <- twostage_admissible_designs(p0, p1, alpha, power, n_max)
nrow(result)
```

```{r admissible-table}
out <- result[, c("n1", "n", "r1", "e1", "r",
                  "alpha_actual", "power_actual",
                  "en_null", "en_alt", "design_type")]
out$alpha_actual <- round(out$alpha_actual, 4)
out$power_actual <- round(out$power_actual, 4)
out$en_null      <- round(out$en_null, 2)
out$en_alt       <- round(out$en_alt, 2)
print(out, row.names = FALSE)
```

## Interpreting the Pareto frontier

Admissibility is defined on three axes: a design survives if no other design
is weakly better on all of $n$, $EN_0$, and $EN_1$ with at least one strict
improvement.

- The **minimax** design minimises the maximum sample size $n$ — useful when
  the budget is fixed and worst-case enrolment is the binding constraint.
- The **optimal null** design minimises $EN_0$ — useful when the null hypothesis
  is expected to hold (the drug is inactive) and early stopping is the priority.
- The **optimal alt** design minimises $EN_1$ — useful when the alternative is
  expected to hold and the priority is small expected size under activity.
- The **optimal mean** design minimises the average $\tfrac{1}{2}(EN_0 + EN_1)$ —
  a compromise that does not commit to either hypothesis being true.
- Other admissible designs represent different trade-offs between the three
  criteria.

```{r pareto-plot, fig.width = 5.5, fig.height = 4, fig.alt = "Scatter plot of EN_null vs EN_alt for admissible designs, with minimax and the three optimal designs labelled"}
plot(result$en_null, result$en_alt,
     xlab = expression(EN[0] ~ "(under null)"),
     ylab = expression(EN[1] ~ "(under alternative)"),
     main = "Pareto frontier of admissible designs",
     pch  = 19, col = "steelblue", cex = 1.2)

highlights <- list(
  c(label = "minimax",      col = "firebrick", pch = "17"),
  c(label = "optimal null", col = "darkgreen", pch = "15"),
  c(label = "optimal alt",  col = "darkorange", pch = "18"),
  c(label = "optimal mean", col = "purple",    pch = "8"))
for (h in highlights) {
  sel <- grepl(h[["label"]], result$design_type, fixed = TRUE)
  points(result$en_null[sel], result$en_alt[sel],
         col = h[["col"]], pch = as.integer(h[["pch"]]), cex = 1.6)
}
legend("topright",
       legend = c("admissible", "minimax", "optimal null",
                  "optimal alt", "optimal mean"),
       pch    = c(19, 17, 15, 18, 8),
       col    = c("steelblue", "firebrick", "darkgreen",
                  "darkorange", "purple"),
       bty    = "n")
```

## Restricting to non-binding designs

Set `non_binding = TRUE` to restrict to designs where the interim efficacy
declaration cannot be overturned at the final analysis ($e_1 \ge r$). This is
the relevant class when the interim efficacy stopping rule is **non-binding**.
The `is_non_binding` column indicates which designs satisfy this constraint.

```{r nonbind-true}
result_nb <- twostage_admissible_designs(p0, p1, alpha, power, n_max,
                                         non_binding = TRUE)
cat(sprintf("%d admissible designs total, %d non-binding\n",
            nrow(result), sum(result$is_non_binding)))
cat(sprintf("%d admissible non-binding designs\n", nrow(result_nb)))
```

## Restricting to Simon two-stage designs

Simon two-stage designs are those with no interim stopping for efficacy —
the trial always proceeds to stage 2 unless stopped early for futility.
In the package's parameterisation this means $e_1 = n_1 + 1$ (the disabled
sentinel value). Pass `simon = TRUE` to restrict the search to this class.

Because Simon designs never make an interim efficacy declaration, the
$e_1 \ge r$ constraint is vacuously satisfied for any $r$, so the
`non_binding` argument has no effect when `simon = TRUE`.

```{r simon}
simon_result <- twostage_admissible_designs(p0, p1, alpha, power, n_max,
                                            simon = TRUE)
nrow(simon_result)
```

All returned designs have $e_1 = n_1 + 1$:

```{r simon-check}
all(simon_result$e1 == simon_result$n1 + 1L)
```

```{r simon-table}
simon_out <- simon_result[, c("n1", "n", "r1", "e1", "r",
                              "alpha_actual", "power_actual",
                              "en_null", "en_alt", "design_type")]
simon_out$alpha_actual <- round(simon_out$alpha_actual, 4)
simon_out$power_actual <- round(simon_out$power_actual, 4)
simon_out$en_null      <- round(simon_out$en_null, 2)
simon_out$en_alt       <- round(simon_out$en_alt, 2)
print(simon_out, row.names = FALSE)
```

## Evaluating a design

`twostage_evaluate_design()` computes the operating characteristics of a
design at an arbitrary vector of response probabilities `p`, returning a
`data.frame` with one row per `p` and columns `p`, `p_success`, `p_futility`,
`p_efficacy1`, and `en`. This is useful both for spot-checking a design
proposed on external grounds (e.g. a published Simon design) and for
plotting OC curves over a grid of `p`.

```{r evaluate-spot}
twostage_evaluate_design(n1 = 15L, n = 25L, r1 = 0L, e1 = 9L, r = 5L,
                         p = c(p0, p1))
```

```{r oc-curve-data}
mm <- result[grepl("minimax", result$design_type), ]
p_grid <- seq(0.01, 0.99, by = 0.01)
oc <- twostage_evaluate_design(mm$n1, mm$n, mm$r1, mm$e1, mm$r, p = p_grid)
head(oc)
```

```{r oc-curve-plot, fig.width = 5.5, fig.height = 4, fig.alt = "OC curve: probability of trial success vs response probability for the minimax design"}
plot(oc$p, oc$p_success,
     type = "l", lwd = 2, col = "firebrick",
     xlab = "Response probability p",
     ylab = "Probability",
     main = "Operating characteristic curve (minimax design)",
     ylim = c(0, 1))
lines(oc$p, oc$p_futility,  lty = 2, lwd = 1.5, col = "steelblue")
lines(oc$p, oc$p_efficacy1, lty = 3, lwd = 1.5, col = "darkgreen")
abline(v = c(p0, p1), col = "grey60", lty = 2)
legend("topleft",
       legend = c("P(success)", "P(futility stop)", "P(interim efficacy)"),
       lty    = c(1, 2, 3), lwd = c(2, 1.5, 1.5),
       col    = c("firebrick", "steelblue", "darkgreen"),
       bty    = "n")
```
