---
title: "Finding Admissible Two-Stage Trial Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Admissible Two-Stage Trial Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(twostage)
```

## Overview

The **twostage** package enumerates admissible two-stage single-arm trial
designs for a binary endpoint. It is aimed at early-phase oncology studies
where a Simon-like design is needed but the full Pareto frontier — not just
the minimax or optimal design — is of interest.

The main entry point is `admissible_designs()`, which takes the four clinical
constraints and returns all Pareto-optimal designs in a single call.

## Design parameters

Each design is defined by five integers $(n_1, n, r_1, e_1, r)$:

| Parameter | Role |
|-----------|------|
| $n_1$ | Stage-1 sample size |
| $n$ | Total sample size ($n_2 = n - n_1$ enrolled in stage 2) |
| $r_1$ | Futility boundary: stop early if $X_1 \le r_1$ ($-1$ = no futility stop) |
| $e_1$ | Interim efficacy boundary: declare success if $X_1 \ge e_1$ ($n_1+1$ = no interim efficacy stop) |
| $r$ | Final success threshold: declare success if $X_1 + X_2 \ge r$ |

By default no irrevocability constraint is applied. Pass `irrevocable = TRUE`
to restrict to designs where $e_1 \ge r$, guaranteeing that once interim
efficacy is declared no stage-2 outcome can overturn it (worst case $X_2 = 0$
still gives $X_1 + X_2 \ge e_1 \ge r$). This is appropriate when the interim
efficacy stopping rule is **non-binding** — i.e. the trial may continue to
stage 2 even after declaring interim success.

## Finding admissible designs

Suppose the null response rate is $p_0 = 0.10$ and the target alternative is
$p_1 = 0.30$. We want at most 5% type I error and at least 80% power, with
total sample size no larger than 40.

```{r constraints}
p0    <- 0.10
p1    <- 0.30
alpha <- 0.05
power <- 0.80
n_max <- 40L
```

`admissible_designs()` performs an exhaustive search and returns the
Pareto-optimal designs on three criteria simultaneously: maximum sample size
($n$), expected sample size under the null ($EN_0$), and expected sample size
under the alternative ($EN_1$). The **minimax** (minimum $n$) and **optimal**
(minimum $EN_0$) designs are flagged in the `design_type` column.

```{r admissible}
result <- admissible_designs(p0, p1, alpha, power, n_max)
nrow(result)
```

```{r admissible-table}
out <- result[, c("n1", "n", "r1", "e1", "r",
                  "alpha_actual", "power_actual",
                  "en_null", "en_alt", "design_type")]
out$alpha_actual <- round(out$alpha_actual, 4)
out$power_actual <- round(out$power_actual, 4)
out$en_null      <- round(out$en_null, 2)
out$en_alt       <- round(out$en_alt, 2)
print(out, row.names = FALSE)
```

## Interpreting the Pareto frontier

Admissibility is defined on three axes: a design survives if no other design
is weakly better on all of $n$, $EN_0$, and $EN_1$ with at least one strict
improvement.

- The **minimax** design minimises the maximum sample size $n$ — useful when
  the budget is fixed and worst-case enrolment is the binding constraint.
- The **optimal** design minimises $EN_0$ — useful when the null hypothesis is
  expected to hold (the drug is inactive) and early stopping is the priority.
- Other admissible designs represent different trade-offs between the three
  criteria.

```{r pareto-plot, fig.width = 5.5, fig.height = 4, fig.alt = "Scatter plot of EN_null vs EN_alt for admissible designs, with minimax and optimal labelled"}
col <- ifelse(result$design_type == "", "steelblue",
       ifelse(result$design_type == "minimax", "firebrick", "darkgreen"))
pch <- ifelse(result$design_type == "", 19, 17)

plot(result$en_null, result$en_alt,
     xlab = expression(EN[0] ~ "(under null)"),
     ylab = expression(EN[1] ~ "(under alternative)"),
     main = "Pareto frontier of admissible designs",
     pch  = pch, col = col, cex = 1.2)
legend("topright",
       legend = c("admissible", "minimax", "optimal"),
       pch    = c(19, 17, 17),
       col    = c("steelblue", "firebrick", "darkgreen"),
       bty    = "n")
```

## Restricting to irrevocable designs

Set `irrevocable = TRUE` to restrict to designs where the interim efficacy
declaration cannot be overturned at the final analysis ($e_1 \ge r$). The
`is_irrevocable` column indicates which designs satisfy this constraint.

```{r irrev-true}
result_irrev <- admissible_designs(p0, p1, alpha, power, n_max,
                                   irrevocable = TRUE)
cat(sprintf("%d admissible designs total, %d irrevocable\n",
            nrow(result), sum(result$is_irrevocable)))
cat(sprintf("%d admissible irrevocable designs\n", nrow(result_irrev)))
```

## Restricting to Simon two-stage designs

Simon two-stage designs are those with no interim stopping for efficacy —
the trial always proceeds to stage 2 unless stopped early for futility.
In the package's parameterisation this means $e_1 = n_1 + 1$ (the disabled
sentinel value). Pass `simon = TRUE` to restrict the search to this class.

Because Simon designs never make an interim efficacy declaration, the
irrevocability constraint ($e_1 \ge r$) is vacuously satisfied for any $r$,
so the `irrevocable` argument has no effect when `simon = TRUE`.

```{r simon}
simon_result <- admissible_designs(p0, p1, alpha, power, n_max, simon = TRUE)
nrow(simon_result)
```

All returned designs have $e_1 = n_1 + 1$:

```{r simon-check}
all(simon_result$e1 == simon_result$n1 + 1L)
```

```{r simon-table}
simon_out <- simon_result[, c("n1", "n", "r1", "e1", "r",
                              "alpha_actual", "power_actual",
                              "en_null", "en_alt", "design_type")]
simon_out$alpha_actual <- round(simon_out$alpha_actual, 4)
simon_out$power_actual <- round(simon_out$power_actual, 4)
simon_out$en_null      <- round(simon_out$en_null, 2)
simon_out$en_alt       <- round(simon_out$en_alt, 2)
print(simon_out, row.names = FALSE)
```

## Evaluating a single design

`evaluate_design()` computes the operating characteristics for one design
without running the full search. This is useful for checking a design
proposed on external grounds (e.g. a published Simon design).

```{r evaluate}
d <- evaluate_design(n1 = 15L, n = 25L, r1 = 0L, e1 = 9L, r = 5L,
                     p0 = p0, p1 = p1)
str(d)
```
