---
title: "Finding Admissible Two-Stage Trial Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Admissible Two-Stage Trial Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(twostage)
```

## Overview

The **twostage** package enumerates admissible two-stage single-arm trial
designs for a binary endpoint. It is aimed at early-phase oncology studies
where a Simon-like design is needed but the full Pareto frontier — not just
the minimax or optimal design — is of interest.

The workflow has three steps:

1. Specify the null/alternative response rates and error constraints.
2. Call `find_feasible_designs()` to enumerate all designs that meet them.
3. Call `find_admissible_designs()` to retain only Pareto-optimal designs.

## Design parameters

Each design is defined by five integers $(n_1, n, r_1, e_1, r)$:

| Parameter | Role |
|-----------|------|
| $n_1$ | Stage-1 sample size |
| $n$ | Total sample size ($n_2 = n - n_1$ enrolled in stage 2) |
| $r_1$ | Futility boundary: stop early if $X_1 \le r_1$ ($-1$ = no futility stop) |
| $e_1$ | Interim efficacy boundary: declare success if $X_1 \ge e_1$ ($n_1+1$ = no interim efficacy stop) |
| $r$ | Final success threshold: declare success if $X_1 + X_2 \ge r$ |

An **irrevocability constraint** is enforced: $e_1 \ge r$. This guarantees
that once interim efficacy is declared, no stage-2 outcome can overturn it
(worst case $X_2 = 0$ still gives $X_1 + X_2 \ge e_1 \ge r$).

## Step 1 — Choose the design constraints

Suppose the null response rate is $p_0 = 0.10$ and the target alternative is
$p_1 = 0.30$. We want at most 5% type I error and at least 80% power, with
total sample size no larger than 40.

```{r constraints}
p0    <- 0.10
p1    <- 0.30
alpha <- 0.05
power <- 0.80
n_max <- 40L
```

## Step 2 — Enumerate feasible designs

`find_feasible_designs()` performs an exhaustive search over all parameter
combinations and returns a data frame of designs that satisfy all constraints.

```{r feasible}
feasible <- find_feasible_designs(p0, p1, alpha, power, n_max)
nrow(feasible)
```

Each row is one design. The key columns are:

```{r feasible-peek}
head(feasible[, c("n1", "n", "r1", "e1", "r",
                  "alpha_actual", "power_actual",
                  "en_null", "en_alt")])
```

## Step 3 — Find admissible designs

Most feasible designs are dominated: there exists another design with a
smaller expected sample size under *both* the null and the alternative.
`find_admissible_designs()` filters to the Pareto frontier on
$(EN_0, EN_1)$.

```{r admissible}
admissible <- find_admissible_designs(feasible)
nrow(admissible)
```

```{r admissible-table}
out <- admissible[, c("n1", "n", "r1", "e1", "r",
                      "alpha_actual", "power_actual",
                      "en_null", "en_alt")]
out$alpha_actual <- round(out$alpha_actual, 4)
out$power_actual <- round(out$power_actual, 4)
out$en_null      <- round(out$en_null, 2)
out$en_alt       <- round(out$en_alt, 2)
print(out, row.names = FALSE)
```

Every admissible design satisfies the irrevocability constraint:

```{r irrevocable}
all(admissible$is_irrevocable)
```

## Interpreting the Pareto frontier

The admissible designs trade off $EN_0$ (expected $n$ when the drug is
inactive — the more common scenario) against $EN_1$ (expected $n$ when the
drug is active). Designs at the top-left of the frontier minimise $EN_0$ at
the cost of a larger $EN_1$; those at the bottom-right do the opposite.

```{r pareto-plot, fig.width = 5.5, fig.height = 4, fig.alt = "Scatter plot of EN_null vs EN_alt for admissible designs"}
plot(admissible$en_null, admissible$en_alt,
     xlab = expression(EN[0] ~ "(under null)"),
     ylab = expression(EN[1] ~ "(under alternative)"),
     main = "Pareto frontier of admissible designs",
     pch  = 19, col = "steelblue")
```

## Evaluating a single design

`evaluate_design()` computes the operating characteristics for one design
without running the full search. This is useful for checking a design
proposed on external grounds (e.g. a published Simon design).

```{r evaluate}
d <- evaluate_design(n1 = 15L, n = 25L, r1 = 0L, e1 = 9L, r = 5L,
                     p0 = p0, p1 = p1)
str(d)
```
